Characterizing Signal Loss in the 21\,cm Reionization Power Spectrum: \\A Revised Study of PAPER-64

### 2018-08-29 PAPER.tex
put half space for 21 cm in title
first pass at major re-structuring. No broken links.

\documentclass[preprint2,numberedappendix,tighten]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
%\documentclass[twocolumn,numberedappendix]{emulateapj}
\shorttitle{Characterizing Signal Loss, Error, and Bias in the 21\,cm Reionization Power Spectrum}
\shortauthors{Cheng et al.}

@@

\begin{document}
\title{Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64}

@@

\begin{abstract}
The Epoch of Reionization (EoR) is an uncharted era in our Universe's history during which the first stars and 
galaxies led to the ionization of neutral hydrogen in the intergalactic medium. There are many experiments investigating the 
EoR by tracing the 21\,cm line of neutral hydrogen, a signal which is very faint and difficult to isolate. With a new generation 
of instruments and a statistical power spectrum detection in our foreseeable future, it has become increasingly important to 
develop techniques that help maximize sensitivity while validating results. Additionally, it is imperative to understand the trade-offs 
between different analysis methods and their effects on power spectrum estimates. In this paper, we focus on three major 
themes --- signal loss, power spectrum error estimation, and bias in measurements. We describe techniques that affect 
these themes using both toy models and data taken by the 64-element configuration of the Donald C. Backer Precision Array 
for Probing the Epoch of Reionization (PAPER). In particular, we highlight how detailed investigations of these themes have 
led to a revised, higher 21\,cm power spectrum upper limit from PAPER-64. While we only focus on a subset of PAPER-64 data in this paper, revised power spectrum limits from the PAPER experiment are presented in a companion paper by Kolopanis et al. (\textit{in prep.}) and supersede results from 
previously published PAPER analyses.
\end{abstract}

@@

By about one billion years after the Big Bang ($z \sim 6$), the first stars and galaxies are thought to have ionized all the 
neutral hydrogen that dominated the baryonic matter content in the Universe. This transition period, during which the first 
luminous structures formed from gravitational collapse and began to emit intense radiation that ionized the cold neutral gas 
into a plasma, is known as the Epoch of Reionization (EoR). The EoR is a relatively unexplored era in our \textit{cosmic dawn}, which spans the birth of the first stars to the full reionization of the intergalactic medium (IGM). Its 
history encodes important information regarding the nature of the first galaxies and the processes of structure formation. 
Direct measurements of the EoR would unlock powerful characteristics about the IGM, revealing connections 
between the matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured 
web of galaxies we observe today (for a review, see \citet{barkana_and_loeb2001}, \citet{furlanetto_et_al2006} and \citet{loeb_furlanetto_2013}).

@@

In addition to the first tentative detection of the EoR signal made by the Experiment to Detect the Global EoR Signature (EDGES; \citealt{bowman_et_al2018}; \citealt{bowman2010}), there are several radio telescope experiments that have succeeded in using 
the 21\,cm signal from hydrogen to place constraints on the brightness of the signal. Examples of experiments investigating the 
mean brightness temperature of the EoR relative to the CMB are the Large Aperture Experiment to Detect the Dark Ages (LEDA; \citealt{bernardi_et_al2016}), the 
Dark Ages Radio Explorer (DARE; \citealt{burns2012}), the Sonda Cosmol\'ogica de las Islas para la Detecci\'on de 
Hidr\'ogeno NeutroSciHi (SCI-HI; \citealt{voytek2014}), the Broadband Instrument for Global HydrOgen ReioNisation Signal 
(BIGHORNS; \citealt{sokolowski2015}), and the Shaped Antenna measurement of the background RAdio Spectrum (SARAS; 
\citealt{patra2015}). Radio interferometers which seek to measure statistical power spectra include the Giant Metre-wave 
Radio Telescope (GMRT; \citealt{paciga_et_al2013}), the LOw Frequency ARray (LOFAR; \citealt{van_haarlem_et_al2013}), 
the Murchison Widefield Array (MWA; \citealt{tingay_et_al2013}), the 21 Centimeter Array (21CMA; 
\citealt{peterson_et_al2004}; \citealt{wu2009}), the Square Kilometre Array (SKA; \citealt{koopmans_et_al2015}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of 
Reionization Array (HERA), which is currently being built, is a next-generation instrument that aims to combine lessons 
learned from previous experiments and is forecast to be able to make a high-significance power spectrum 
detection with an eventual $350$ elements using current analysis techniques (\citealt{pober_et_al2014}; \citealt{liu_parsons_2016}; \citealt{dillon_parsons2016}; \citealt{deboer_et_al2017}).

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013b, bernardi_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three topics (signal loss, error estimation, and bias) that are essential to investigate 
for a robust 21\,cm power spectrum analysis. We also highlight four power spectrum techniques (fringe-rate filtering, weighting, bootstrapping, jackknife testing) and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER. In this study we use a subset of PAPER-64 data to illustrate our revised analysis methods, while a companion paper, Kolopanis et al. (\textit{in prep.}), builds off of the methods in this paper to present revised PAPER-64 results for multiple redshifts and baseline types.

Finally, this paper adds to the growing foundations of lessons which have been documented, for example, in \citet{Paciga2013}, \citet{Patil2016}, and \citet{Jacobs2016}, by the GMRT, LOFAR, and MWA projects respectively. These lessons are imperative as the community as a whole moves towards higher sensitivities and potential EoR detections.

This paper is organized into two main halves. In Section \ref{sec:Themes} we introduce the three themes of our focus, using 
toy models to develop intuition for each one. In Section \ref{sec:CaseStudy} we present a case study into each theme using data 
from the PAPER-64 array, highlighting key changes from the methods used in the previously published result in \citet{ali_et_al2015}, henceforth known as \citetalias{ali_et_al2015}, which have led to a 
revised PAPER-64 power spectrum result (Kolopanis et al. (\textit{in prep.})). We conclude in Section \ref{sec:Con}.

% SECTION 2 INTRO ---------------------------------------------------------------------------------

\section{Power Spectrum Themes and Techniques}
\label{sec:Themes}

There are many choices a 21\,cm data analyst must consider. %For PAPER, its grid configuration allows the use of many signal processing tricks to down-weight foregrounds without reference to a sky model. However, even with the advantage of being agnostic about the sky, many questions still remain.
%, though the space is significantly constrained by 
%the type of instrument being used.  The PAPER instrument in its grid configuration focuses most sensitivity on 
%only a few baselines but imaging performance is poor. In this space, signal processing tricks which filter or 
%down-weight foreground modes without reference to a sky model have the advantage of being agnostic about 
%the sky. However, even within this problem sub-space, questions remain.
How can time-ordered measurements be combined? How can the 
variance of the data be estimated? In what way(s) can the data be weighted to suppress contaminated modes while not 
destroying an EoR signal? How can a statistically significant detection of a signal be properly identified? Many common techniques, such as 
averaging data, weighting, bootstrapping, and jackknife testing, address these issues but harbor additional trade-offs. For 
example, an aggressive filtering method may succeed in eliminating interfering systematics but comes at the cost of losing 
some EoR signal. A chosen weighting scheme may theoretically maximize sensitivity but fail to suppress foregrounds in practice.

Though there are many data analysis choices, measuring the statistical 21\,cm power spectrum ultimately requires robust 
methods for determining accurate confidence intervals and rigorous techniques to identify and control systematics.  In this 
paper, we focus on three 21\,cm power spectrum themes that encapsulate this goal and discuss four techniques that interplay 
with each other and impact the themes. We will give brief definitions now, and build intuition for each theme in the sections to 
follow.

%\begin{center}
%Power Spectrum Themes
%\end{center}

%A deep understanding of the following three themes is essential for the accuracy and interpretation of a 21\,cm power spectrum result. Stemming from a re-analysis of PAPER-64 data, we believe these themes serve as an important checklist for a rigorous power spectrum analysis.
\begin{itemize}
\item \textbf{Signal Loss} (Section \ref{sec:SiglossOverview}): Signal loss refers to attenuation of the target \textit{cosmological} signal 
in a power spectrum estimate. Certain analysis techniques can cause this loss, and if the amount of loss is not quantified accurately, it could lead to false non-detections and overly aggressive upper limits. Determining whether an analysis pipeline is lossy, and estimating the amount of loss if so, has subtle challenges but is necessary to ensure the accuracy of any result. 
\item \textbf{Error Estimation} (Section \ref{sec:ErrorOverview}): Confidence intervals on a 21\,cm power spectrum result 
determine the difference between a detection and a null result, which have two very different implications. Additionally, accurate error estimation is crucial for the comparison of results to theoretical models. Errors can be 
estimated in a variety of ways, and we will discuss a few of them.
\item \textbf{Bias} (Section \ref{sec:BiasOverview}): There are several possible sources of power offset in a visibility 
measurement that can show up as a detection in a power spectrum, such as bias from noise and foregrounds. In particular, a 
successful EoR detection would also imitate a bias. Proving that a bias is an EoR detection may be the most difficult challenge for future 21\,cm 
analyses, as it is crucial to be able to distinguish a detection of foreground leakage, for example, from that of EoR. In this paper 
we will highlight some sources of bias, discuss ways to mitigate their effects, and describe example tests that a true EoR detection must 
pass.
\end{itemize}

%\begin{center}
%Power Spectrum Techniques
%\end{center}

The following techniques each have advantages when it comes to maximizing sensitivity and understanding systematics in 
data. However, some have limitations, and we will discuss circumstances in which there are trade-offs. We choose to focus on 
these four techniques because they represent major steps in PAPER's power spectrum pipeline, with several of them also 
being standard steps in general 21\,cm analyses.
\begin{itemize}
\item \textbf{Fringe-rate filtering:} Fringe-rate filtering is an averaging scheme for time-ordered data 
(\citealt{parsons_et_al2016}). Broadly, a fringe-rate filter averages visibilities in time to produce a smaller number of more sensitive 
independent samples. However, such a filter also affects the presence 
of foregrounds and systematics. We explain the trade-offs of filtering in more detail in Section \ref{sec:toymodel_frf}.
\item \textbf{Weighting:} A dataset can be weighted to emphasize certain features and minimize others. One particular flavor of 
weighting employed by previous PAPER analyses is inverse covariance weighting in frequency, which is a generalized version of inverse variance 
weighting that also takes into account frequency correlations (\citealt{liu_tegmark2011}; \citealt{dillon_et_al2013a}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{dillon_et_al2014}; \citealt{dillon_et_al2015}). Using such a technique enables the down-weighting of contaminant modes that obey a different covariance structure from that of cosmological modes. However, a challenge of inverse covariance 
weighting is in estimating a covariance matrix that is closest to the true covariance of the data; the discrepancy between the two can have large impacts on signal loss. We investigate the impact of different types of weighting on signal loss in Section \ref{sec:SiglossOverview}.
\item \textbf{Bootstrapping:} In addition to using theoretical models for covariance matrices and theoretical error estimation 
methods, bootstrapping is one way to estimate errors. Namely, bootstrapping is a useful method for estimating errors of a dataset from 
itself (\citealt{andrae2010}). By randomly drawing many subsamples of the data, we obtain a sense of its inherent variance, though there are subtleties to 
consider such as the independence of values in a dataset. We explore this potential pitfall of bootstrapping in Section \ref{sec:ErrorOverview}.
\item \textbf{Jackknife testing:} A resampling technique useful for estimating bias, jackknives can be taken along different 
dimensions of a dataset to cross-validate results. In particular, null tests can be used to verify whether results are free of 
systematics, as done with CO power spectra (\citealt{keating_et_al2016}) and CMB measurements (see e.g. \citealt{ade_et_al2008}; \citealt{chiang_et_al2010}; \citealt{bischoff_et_al2011}; \citealt{das_et_al2011b}; \citealt{araujo_et_al2012}; \citealt{crites_et_al2015}; \citealt{ade_et_al2016}; \citealt{ade_et_al2017}; \citealt{sherwin_et_al2017}). An EoR detection must pass both jackknife and null tests, which we highlight in Section \ref{sec:JackknifeOverview}.
\end{itemize}

In the next three subsections, we study each theme in depth, focusing on how power spectrum technique trade-offs affect each. 
We use toy data models to develop intuition into why certain analysis choices may be appealing and discuss ways in which 
they are limited. We highlight problems that can arise regarding each theme and offer suggestions to mitigate the issues. 
Ultimately, we show that rigorous investigations into signal loss, error estimation, and bias must be performed for robust 21\,cm results.

% SECTION 2 SIGLOSS ---------------------------------------------------------------------------------
%\color{blue}

\subsection{Signal Loss}
\label{sec:SiglossOverview}
% JEA - We never totally nail down what ``signal loss'' is ...

Signal loss can arise in a variety of ways in an analysis pipeline, such as by fitting a polynomial during spectral calibration, applying a delay domain filter, or deriving weights from data and applying them to itself. Here we focus on signal loss associated with 
% JEA - we're not considering a general problem, but are really just focused on the empirical QE and its discontents, so let's say that
the use of an empirically estimated covariance matrix with the ``optimal quadratic estimator'' formalism.
This loss was significantly 
%applying a weighting matrix to data, a loss that can be significant depending on the choice of weighting and one that was previously 
under-estimated in the \citetalias{ali_et_al2015} analysis.

@@

\subsubsection{The Quadratic Estimator Method}
\label{sec:QE}

@@

The goal of power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission. Prior to power spectrum estimation, the data will often have been prepared to have minimal foregrounds by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual of some subtraction process not in the power spectrum domain. If an accurate estimate of the total covariance of the data is known, including both the desired signal and any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the desired signal, as shown in 
% JEA - I cut this b/c I wanted to step back and explain a bit more clearly what we were actually hoping for 
%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. 
%Before showing how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER analysis.
We provide a brief summary of the optimal quadratic estimator and explain its advantages below.

% JEA - The discussion of dimensions here is confusing ... "a baseline" is mentioned, but it's not clear if the redundant baseline averaging has already happened or what.  I believe we are only considering covariance between frequencies, but this isn't explicitly said, and the matrix equations don't make sense unless all the dimensions are N_f N_t.  
%We begin with our 
The measured visibilities for a single baseline in Jy are arranged as a data vector, $\textbf{x}$. It has length $N_{t} N_{f}$,
% JEA - not sure this helps
%, but in practice we manipulate it as an array with dimensions ($N_{f}, N_{t}$), 
where $N_{t}$ is the number of time integrations and $N_{f}$ is the number of frequency channels. The covariance of the data is given by 
\begin{equation}
\textbf{C} \equiv \langle\textbf{xx}^{\dagger}\rangle = \textbf{S} + \textbf{U}
\end{equation}
where the average over an ensemble of data realizations produces the true covariance, and we further assume it may be written as the sum of the desired cosmological signal $\textbf{S}$ and other terms $\textbf{U}$.

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $P^\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $P^\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = \sum_\alpha P^\alpha \frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \sum_\alpha P^\alpha \textbf{Q}^\alpha
%\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we define $\frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \textbf{Q}^{\alpha}$.
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{P}^{\alpha}  = \sum_\beta ({\F^{-1}})^{\alpha\beta} (\widehat{q}^{\beta} - \widehat{b}^{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \textbf{Q}^{\alpha} \C^{-1} \textbf{Q}^{\beta} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\widehat{q}^\alpha =  \half \textbf{x}^\dagger \invC \textbf{Q}^{\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}^{\alpha} = \half \Tr\left( \mathbf{U} \invC \textbf{Q}^{\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{P}^{\alpha} \rangle = P^\alpha
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}.

It is clear that in order to obtain an unbiased estimate in the presence of contaminants, the estimator defined by Equation \eqref{eq:OQE} must be capable of ``suppressing'' or ``removing'' their effects. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

>>>>

\documentclass[preprint2,numberedappendix,tighten]{aastex6}  % USE THIS TO MAKE BIB, THEN FORMAT USING EMULATEAPJ
%\documentclass[twocolumn,numberedappendix]{emulateapj}
\shorttitle{Characterizing Signal Loss in the 21\,cm Reionization Power Spectrum}
\shortauthors{Cheng et al.}

@@

\begin{document}
\title{Characterizing Signal Loss in the 21\,cm Reionization Power Spectrum: \\A Revised Study of PAPER-64}

@@

\begin{abstract}
The Epoch of Reionization (EoR) is an uncharted era in our Universe's history during which the first stars and 
galaxies led to the ionization of neutral hydrogen in the intergalactic medium. There are many experiments investigating the 
EoR by tracing the 21\,cm line of neutral hydrogen. Because this signal is very faint and difficult to isolate, it is crucial to develop analysis techniques that maximize sensitivity and suppress contaminants in data. It is also imperative to understand the trade-offs between different analysis methods and their effects on power spectrum estimates. Specifically, with a new generation of instruments and a statistical power spectrum detection in our foreseeable future, it has become increasingly important to understand how certain analysis choices can lead to the loss of the EoR signal. In this paper, we focus on signal loss specifically associated with quadratic estimators and power spectrum estimation. We describe the origin of this loss using both toy models and data taken by the 64-element configuration of the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER). In particular, we highlight how detailed investigations of signal loss have led to a revised, higher 21\,cm power spectrum upper limit from PAPER-64. Additionally, we summarize errors associated with power spectrum error estimation that were previously unaccounted for. While we only focus on a subset of PAPER-64 data in this paper, revised power spectrum limits from the PAPER experiment are presented in a forthcoming paper by Kolopanis et al. (\textit{in prep.}) and supersede results from previously published PAPER analyses.
\end{abstract}

@@

By about one billion years after the Big Bang ($z \sim 6$), the first stars and galaxies are thought to have ionized all the 
neutral hydrogen that dominated the baryonic matter content in the Universe. This transition period, during which the first 
luminous structures formed from gravitational collapse and began to emit intense radiation that ionized the cold neutral gas 
into a plasma, is known as the Epoch of Reionization (EoR). The EoR is a relatively unexplored era in our Universe's history, which spans the birth of the first stars to the full reionization of the intergalactic medium (IGM). This epoch encodes important information regarding the nature of the first galaxies and the processes of structure formation. 
Direct measurements of the EoR would unlock powerful characteristics about the IGM, revealing connections 
between the matter distribution exhibited via cosmic microwave background (CMB) studies and the highly structured 
web of galaxies we observe today (for a review, see \citet{barkana_and_loeb2001}, \citet{furlanetto_et_al2006} and \citet{loeb_furlanetto_2013}).

@@

In addition to the first tentative detection of our \textit{Cosmic Dawn} (pre-reionization) made by the Experiment to Detect the Global EoR Signature (EDGES; \citealt{bowman_et_al2018}; \citealt{bowman2010}), there are several radio telescope experiments that have succeeded in using 
the 21\,cm signal from hydrogen to place constraints on the brightness of the signal. Examples of experiments investigating the 
mean brightness temperature of the 21\,cm signal relative to the CMB are the Large Aperture Experiment to Detect the Dark Ages (LEDA; \citealt{bernardi_et_al2016}), the 
Dark Ages Radio Explorer (DARE; \citealt{burns2012}), the Sonda Cosmol\'ogica de las Islas para la Detecci\'on de 
Hidr\'ogeno NeutroSciHi (SCI-HI; \citealt{voytek2014}), the Broadband Instrument for Global HydrOgen ReioNisation Signal 
(BIGHORNS; \citealt{sokolowski2015}), and the Shaped Antenna measurement of the background RAdio Spectrum (SARAS; 
\citealt{patra2015}). Radio interferometers which seek to measure statistical power spectra include the Giant Metre-wave 
Radio Telescope (GMRT; \citealt{paciga_et_al2013}), the LOw Frequency ARray (LOFAR; \citealt{van_haarlem_et_al2013}), 
the Murchison Widefield Array (MWA; \citealt{tingay_et_al2013}), the 21 Centimeter Array (21CMA; 
\citealt{peterson_et_al2004}; \citealt{wu2009}), the Square Kilometre Array (SKA; \citealt{koopmans_et_al2015}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of 
Reionization Array (HERA), which is currently being built, is a next-generation instrument that aims to combine lessons 
learned from previous experiments and has a forecasted sensitivity capable of a high-significance power spectrum 
detection with an eventual $350$ elements using current analysis techniques (\citealt{pober_et_al2014}; \citealt{liu_parsons_2016}; \citealt{dillon_parsons2016}; \citealt{deboer_et_al2017}).

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013b, bernardi_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of how data analysis choices, which are often chosen to maximize sensitivity and minimize contaminants, affect power spectrum results. More specifically, it is imperative to develop techniques that ensure 
the accurate extraction and recovery of the EoR signal, despite the analysis method chosen and how much loss may have resulted. In this paper, we discuss signal loss associated with specifically with power spectrum estimation --- a topic that is essential to investigate for a robust 21\,cm power spectrum analysis and one that has been underestimated in previous PAPER analyses. We first approach this topic from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER. In this study we use a subset of PAPER-64 data to illustrate our revised analysis methods, while a related paper, Kolopanis et al. (\textit{in prep.}), builds off of the methods in this paper to present revised PAPER-64 results for multiple redshifts and baseline types.

Finally, we also highlight several additional errors made in previous PAPER analyses, including those related to bootstrapping and error estimation. This paper adds to the growing foundations of lessons which have been documented, for example, in \citet{Paciga2013}, \citet{Patil2016}, and \citet{Jacobs2016}, by the GMRT, LOFAR, and MWA projects respectively. These lessons are imperative as the community as a whole moves towards higher sensitivities and potential EoR detections.

This paper is organized into three main sections. In Section \ref{sec:SiglossOverview} we use toy models to develop intuition about signal loss, its origin, and its subtleties. In Section \ref{sec:CaseStudy}, we present a case study using data from the PAPER-64 array, highlighting key changes from the signal loss methods used in the published result in \citet{ali_et_al2015}, henceforth known as \citetalias{ali_et_al2015}, which previously underestimated loss. In Section \ref{sec:OtherErrors}, we summarize additional lessons learned since \citetalias{ali_et_al2015} that have shaped our revised analysis. Finally, we conclude in Section \ref{sec:Con}.

% SECTION 2 SIGLOSS TOY MODEL ---------------------------------------------------------------------------------
%\color{blue}

\section{Signal Loss Toy Models}
\label{sec:SiglossOverview}
% JEA - We never totally nail down what ``signal loss'' is ...

Signal loss refers to attenuation of the target \textit{cosmological} signal 
in a power spectrum estimate. Certain analysis techniques can cause this loss, and if the amount of loss is not quantified accurately, it could lead to false non-detections and overly aggressive upper limits. Determining whether an analysis pipeline is lossy, and estimating the amount of loss if so, has subtle challenges but is necessary to ensure the accuracy of any result.

Signal loss can arise in a variety of ways in an analysis pipeline, such as by fitting a polynomial during spectral calibration, applying a delay domain filter, or deriving weights from data and applying them to itself. Here we focus on signal loss associated with 
% JEA - we're not considering a general problem, but are really just focused on the empirical QE and its discontents, so let's say that
the latter.

A dataset can be weighted to emphasize certain features and minimize others, and one particular flavor of weighting employed by previous PAPER analyses is inverse covariance weighting in frequency, which is a generalized version of inverse variance weighting that also takes into account frequency correlations (\citealt{liu_tegmark2011}; \citealt{dillon_et_al2013a}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{dillon_et_al2014}; \citealt{dillon_et_al2015}). Using such a technique enables the down-weighting of contaminant modes that obey a different covariance structure from that of cosmological modes. However, a challenge of inverse covariance 
weighting is in estimating a covariance matrix that is closest to the true covariance of the data; the discrepancy between the two, as we will see, can have large impacts on signal loss. Here we focus specifically on loss associated with the use of an empirically estimated covariance matrix with the ``optimal quadratic estimator'' formalism.
This loss was significantly 
%applying a weighting matrix to data, a loss that can be significant depending on the choice of weighting and one that was previously 
underestimated in the \citetalias{ali_et_al2015} analysis.

@@

\subsection{The Quadratic Estimator Method}
\label{sec:QE}

@@

We begin with an overview of the quadratic estimator (QE) formalism used for power spectrum estimation. The goal of power spectrum analysis is to produce an unbiased estimator of the EoR power spectrum in the presence of both noise and foreground emission. Prior to power spectrum estimation, the data will often have been prepared to have minimal foregrounds by some method of subtraction, so this foreground emission may appear either directly (because it was not subtracted) or as a residual of some subtraction process not in the power spectrum domain. If an accurate estimate of the total covariance of the data is known, including both the desired signal and any contaminants, then the ``optimal quadratic estimator'' formalism provides a method of producing a minimum variance, unbiased estimator of the desired signal, as shown in 
% JEA - I cut this b/c I wanted to step back and explain a bit more clearly what we were actually hoping for 
%Driven by the need to mitigate foreground bias, PAPER's previous analyses use a weighting method that aims to down-weight foregrounds. This weighting is applied to data, which is then propagated into a final estimator using the power spectrum estimation technique of quadratic estimators (QE) as done in 
\citet{liu_tegmark2011}, \citet{dillon_et_al2013a}, \citet{liu_et_al2014a}, \citet{liu_et_al2014b}, \citet{trott_et_al2012}, \citet{dillon_et_al2014}, \citet{dillon_et_al2015}, \citet{switzer_et_al2015}, and \citet{trott_et_al2016}. 
%Before showing how signal loss can arise when using different weighting matrices, we first summarize QE as performed in the PAPER analysis.

% JEA - The discussion of dimensions here is confusing ... "a baseline" is mentioned, but it's not clear if the redundant baseline averaging has already happened or what.  I believe we are only considering covariance between frequencies, but this isn't explicitly said, and the matrix equations don't make sense unless all the dimensions are N_f N_t.  
%We begin with our 
Suppose that the measured visibilities for a single baseline in Jy are arranged as a data vector, $\textbf{x}$. It has length $N_{t} N_{f}$,
% JEA - not sure this helps
%, but in practice we manipulate it as an array with dimensions ($N_{f}, N_{t}$), 
where $N_{t}$ is the number of time integrations and $N_{f}$ is the number of frequency channels. The covariance of the data is given by 
\begin{equation}
\textbf{C} \equiv \langle\textbf{xx}^{\dagger}\rangle = \textbf{S} + \textbf{U}
\end{equation}
where the average over an ensemble of data realizations produces the true covariance, and we further assume it may be written as the sum of the desired cosmological signal $\textbf{S}$ and other terms $\textbf{U}$.

We are interested in estimating the three-dimensional power spectrum of the EoR.  
Visibilities are measurements of the Fourier transform of the sky along two spatial dimensions (using the flat-sky approximation), and since we are interested in three-dimensional Fourier modes we only need to take one Fourier transform of our visibilities along the line-of-sight dimension.  We consider band powers $P^\alpha$ of the power spectrum of $\textbf{x}$ over some range in cosmological $\mathbf{k}$, where $\alpha$ indexes a waveband in $k_{\parallel}$ (a cosmological wavenumber $k_{\parallel}$ is the Fourier dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}), which is a good approximation for the short baselines that PAPER analyzes).  The fundamental dependence of the covariance on the power spectrum band powers $P^\alpha$ is encoded as 
% JEA - Should distinguish slightly more clearly between Q and dC/dp_alpha
\begin{equation}
\textbf{S} = \sum_\alpha P^\alpha \frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \sum_\alpha P^\alpha \textbf{Q}^\alpha
%\textbf{S} = p_{\alpha} \frac{\partial\textbf{C}}{\partial p_{\alpha}}
%\frac{\partial\textbf{C}}{\partial p_{\alpha}} \equiv \textbf{Q}^{\alpha}
\end{equation}
where we define $\frac{\partial\textbf{C}}{\partial P^\alpha} \equiv \textbf{Q}^{\alpha}$. In other words, $\textbf{Q}$ describes the response of the covariance to a change in the power spectrum, relating a quadratic statistic of the data (the covariance) to a quadratic statistic in Fourier-space (the power spectrum). 
% JEA - ah, the dimensions again
% which has dimensions ($N_{f}$,$N_{f}$).

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{P}^{\alpha}  = \sum_\beta ({\F^{-1}})^{\alpha\beta} (\widehat{q}^{\beta} - \widehat{b}^{\beta} )
\end{equation}
where $\F$ is the Fisher matrix (which determines errors on the power spectrum estimate)
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \textbf{Q}^{\alpha} \C^{-1} \textbf{Q}^{\beta} \right),
\end{equation}
$\widehat{q}$ is the un-normalized power spectrum estimate
\begin{equation}
\label{eq:OQEQuadratic}
\widehat{q}^\alpha =  \half \textbf{x}^\dagger \invC \textbf{Q}^{\alpha}  \invC \textbf{x},
\end{equation}
and $\widehat{b}$ is the additive bias
\begin{equation}
\label{eq:OQELinear}
\widehat{b}^{\alpha} = \half \Tr\left( \mathbf{U} \invC \textbf{Q}^{\alpha} \invC \right).
\end{equation}
The power spectrum estimator in Equation \eqref{eq:OQE} is the minimum variance (smallest error bar) estimate of the power spectrum subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{P}^{\alpha} \rangle = P^\alpha
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}.

Intuitively, the estimator must be capable of ``suppressing" or ``removing" the effects of contaminates in order to obtain an unbiased estimate of the power spectrum. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.
